\documentclass{beamer}

\title{6 Degrees of Wikipedia}
\author{Troy Astorino, Neil Forrester}
\date{April 10, 2013}
\institute[6.834 -- MIT]{Cognitive Robotics \\ Massachusetts Institute of Technology}

\usepackage{graphicx}
\usepackage{amsmath}

\usetheme{CambridgeUS}
\usecolortheme{beaver}
\setbeamertemplate{section in toc}[square]
%\setbeamercolor{section number projected}[bg=black, fg=red]
\setbeamertemplate{navigation symbols}{} % remove navigation symbols

\begin{document}
\begin{frame}
  \maketitle
\end{frame}

\begin{frame}
  \title{The 6 degrees of Wikipedia Game}

% image of Chewbacca to Tussock (grass)
\end{frame}

\begin{frame}
  \title{Connection to co-occurences}
% Constraints on the problem make fetching expensive
% high cost action makes similar situation to co-occurences
\end{frame}

\begin{frame}
  \title{Category Graph as Knowledge Graph}
% all wikipedia articles in categories (less categories than articles)
% use categories to inform search
\end{frame}

\begin{frame}
  \title{Probability that an article is in a category}

\[P(t_i \in c_j | t_i = [w_1, w_2, \dots, w_n]) = \frac{P(t_i = [w_1, w_2,
  \dots, w_n] | t_i \in c_j) P(t_i \in c_j)}{P(t_i = [w_1, w_2, \dots, w_n])}\]

\end{frame}

\begin{frame}
\title{Assume independence of words}

\begin{multline*}
P(t_i = [w_1, w_2, \dots, w_n] | t_i \in c_j) \\
\begin{aligned}
& = P(w_1 \in t_i \cap w_2 \in t_i \cap \dots \cap w_n \in t_i | t_i \in c_j) \\
& \approx P(w_1 \in t_i | t_i \in c_j) P(w_2 \in t_i | t_i \in c_j) \cdots P(w_n
\in t_i | t_i \in c_j) \\
\end{aligned}
\end{multline*}

\end{frame}

\begin{frame}
\title{Training data}

\[P(w_l \in t | t \in c_j) \propto \displaystyle \sum_{t \in c_j} \sum_l 
    \delta (w_l \in t)\]
such that
\[\displaystyle \sum_l P(w_l \in t | t \in c_j) = 1\]

\vspace{2em}

\[P(t_i \in c_j) = \frac{\text{\# of articles in } c_j}{\text{\# of articles}}\]

\end{frame}


\begin{frame}
\frametitle{Search strategy}

% how we are using category graph path and generality of categories to inform search
\end{frame}

\begin{frame}
\frametitle{Limiting scope of Wikipedia}

\begin{itemize}
\item Consider only the $x$ most viewed articles. Only include categories that
  these articles are in.

\item Consider only the top $x$ categories ranked by number of pages in the
  category. Only include pages in those categories.

\item Consider only the top $x$ categories ranked by number of subcategories in the
  category. Only include pages in those categories.

\item Consider only articles with more than $x$ words. Only include categories
  those pages are in.
\end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Current Status}
\end{frame}
\end{document}